Uncertainty Quantification and Inference of Manning's Friction Coefficient using DART Buoys Data during \tohoku Tsunami

### 2014-01-14 paper/conc.tex
fixed figures and removed comments from paper

The present study aimed at estimating Manning's $n$ friction coefficient 
which is crucial in the accurate prediction of water surface elevation in 
tsunami events modeling. We proposed a three-dimensional representation
of Manning's $n$ friction coefficient that was characterized using iso-baths 
to define three distinct regions in the domain and represented each region
with a single Manning's $n$ coefficient. The three regions were as follows:
an on-shore region (initially above sea-level), a near-shore region (between
sea-level and the 200 meter iso-bath), and the deep-water region (deeper than
200 meters) whose Manning's $n$ friction coefficient were denoted by $n_1$, $n_2$, 
and $n_3$ respectively. The estimation relied on a Bayesian
inference approach that sharpens the initial estimates of the three uncertain parameters
using real observations. In our case, we used the DART buoys system data that provides water 
surface elevation data. We specifically used the data collected at four different gauges 
during \tohoku tsunami. To accelerate the Bayesian inference, we relied on the polynomial 
chaos expansions to produce a faithful and efficient surrogate of the forward model \geoclaw. 
The main results of this work were:
\begin{itemize}

>>>>

The present study aimed at estimating Manning's $n$ friction coefficient 
which is crucial in the accurate prediction of water surface elevation in 
tsunami events modeling. We proposed a three-parameter representation
of Manning's $n$ friction coefficient that was characterized using iso-baths 
to define three distinct regions in the domain and represented each region
with a single Manning's $n$ coefficient. The three regions were as follows:
an on-shore region (initially above sea-level), a near-shore region (between
sea-level and the 200 meter iso-bath), and the deep-water region (deeper than
200 meters) whose Manning's $n$ friction coefficient were denoted by $n_1$, $n_2$, 
and $n_3$ respectively. The estimation relied on a Bayesian
inference approach that sharpens the initial estimates of the three uncertain parameters
using real observations. In our case, we used the DART buoys system data that provides water 
surface elevation data. We specifically used the data collected at four different gauges 
during \tohoku tsunami. To accelerate the Bayesian inference, we relied on the polynomial 
chaos expansions to produce a faithful and efficient surrogate of the forward model \geoclaw. 
The main results of this work were:
\begin{itemize}

### 2014-01-14 paper/intro.tex
fixed figures and removed comments from paper

In this work, we focus our attention on the uncertainty in the friction
parameterization used in tsunami simulations. We present a \alert{three}-dimensional 
representation of Manning's $n$ coefficient and an efficient method
to estimate this representation using real water elevation data. The method proposed here 
is based on recent developments in Uncertainty Quantification (UQ) methods that allow 
probing of the sensitivity of realistic tsunami  models to uncertain parameters
without modifying the forward model and inferring those parameters
from a number of observations.  In particular, we implement an inverse modeling
approach that relies on the Bayesian inference technique.  We resort to
Polynomial Chaos (PC) expansions to construct a faithful surrogate model of the
response of the large and complex geophysical model simulations \cite{Najm:2009,Alexanderian2012,sraj:2013a,sraj:2013b}.
The construction of the surrogate enables us to efficiently implement the Bayesian
inference technique as it dramatically reduces the cost of sampling the
posterior distribution.

@@

In this work, we follow a similar procedure as in \cite{sraj:2013a,sraj:2013b}
to estimate Manning's $n$ coefficient using the \tohoku tsunami
event as a case-study. The quantity of interest is taken to be the surface
elevation $\eta$ and is compared to the available NOAA Deep-ocean Assessment and
Reporting Tsunamis (DART) buoys.  The forward model runs are performed using the
\geoclaw package following the setup in MacInnes \emph{et al.}~\cite{MacInnes:2013cr} 
including the two source models analyzed there.

The structure of this manuscript is as follows. First, we discuss the uncertainties 
of the Manning'g $n$ coefficient to motivate the UQ problem in Section~\ref{sec:manning}
and propose a \alert{three}-dimensional representation of $n$. 
Section~\ref{sec:tohoku} describes our case-study, the \tohoku tsunami,
summarize the forward simulations setup, and present the DART buoys observational data. 
Section~\ref{sec:formu} presents the two main methods employed in the UQ analysis:
Bayesian inference that brings together observations 
and model results (Section~\ref{sec:inference}) 
and PC expansions that are used to build the surrogate model (Section~\ref{sec:uqpce}). Section~\ref{sec:results} describes the main results in 
three subsections: first, we perform error and convergence studies
as evidences that the PC surrogate is a faithful 
representation of the \geoclaw model; second, we use the PC surrogate to 
perform a statistical analysis and compute the  sensitivity of simulated water elevations
to the uncertain input parameters; and third, we present the results of the
posterior distributions obtained using Bayesian inference by MCMC. A discussion of our findings and methodology is finally presented in Section~\ref{sec:conc}.

\comment{
\begin{itemize}
    \item There are two places where we refer to the parameterization as three-dimensional, I am not sure if this is really three-dimensional.  I know that we are using depth contours but the parameterization is two-dimensional.  I have a feeling that readers would take issue with this.
    \item Ihab: the representation is 3D meaning we used three different values of n in three regions. The same concept was presented in Mayo paper where they did 2D representation.
    \item I think the order here needs to be revised to talk about the forward model and the equations we are solving first since the Manning's $n$ discussion refers to equations there. 
    \item Ihab: Our main objective is to show the UQ techniques and how they work. The equations are known and we do not need to stress on them. We can even remove the whole equation or forward model section.
\end{itemize}
}

>>>>

In this work, we focus our attention on the uncertainty in the friction
parameterization used in tsunami simulations. We present a three-parameter 
representation of Manning's $n$ coefficient and an efficient method
to estimate this representation using real water elevation data. The method proposed here 
is based on recent developments in Uncertainty Quantification (UQ) methods that allow 
probing of the sensitivity of realistic tsunami  models to uncertain parameters
without modifying the forward model and inferring those parameters
from a number of observations.  In particular, we implement an inverse modeling
approach that relies on the Bayesian inference technique.  We resort to
Polynomial Chaos (PC) expansions to construct a faithful surrogate model of the
response of the large and complex geophysical model simulations \cite{Najm:2009,Alexanderian2012,sraj:2013a,sraj:2013b}. The construction of the surrogate enables us to efficiently implement the Bayesian
inference technique as it dramatically reduces the cost of sampling the
posterior distribution.

@@

In this work, we follow a similar procedure as in \cite{sraj:2013a,sraj:2013b}
to estimate Manning's $n$ coefficient using the \tohoku tsunami
event as a case-study. The quantity of interest is taken to be the water surface
elevation and is compared to the available NOAA Deep-ocean Assessment and
Reporting Tsunamis (DART) buoys.  The forward model runs are performed using the
\geoclaw package following the setup in MacInnes \emph{et al.}~\cite{MacInnes:2013cr} 
including the two source models analyzed there.

The structure of this manuscript is as follows. First, we discuss the uncertainties 
of the Manning'g $n$ coefficient to motivate the UQ problem in Section~\ref{sec:manning}
and propose a three-parameter representation of $n$. 
Section~\ref{sec:tohoku} describes our case-study, the \tohoku tsunami,
summarizes the forward simulations setup, and presents the DART buoys observational data. 
Section~\ref{sec:formu} presents the two main methods employed in the UQ analysis:
Bayesian inference that brings together observations 
and model results (Section~\ref{sec:inference}) 
and PC expansions that are used to build the surrogate model (Section~\ref{sec:uqpce}). Section~\ref{sec:results} describes the main results in 
three subsections: first, we perform error and convergence studies
as evidences that the PC surrogate is a faithful 
representation of the \geoclaw model; second, we use the PC surrogate to 
perform a statistical analysis and compute the  sensitivity of simulated water elevations
to the uncertain input parameters; and third, we present the results of the
posterior distributions obtained using Bayesian inference by MCMC. A discussion of our findings and methodology is finally presented in Section~\ref{sec:conc}.

%\comment{
%\begin{itemize}
%    \item There are two places where we refer to the parameterization as three-dimensional, I am not sure if this is really three-dimensional.  I know that we are using depth contours but the parameterization is two-dimensional.  I have a feeling that readers would take issue with this.
%    \item Ihab: the representation is 3D meaning we used three different values of n in three regions. The same concept was presented in Mayo paper where they did 2D representation.
%    \item I think the order here needs to be revised to talk about the forward model and the equations we are solving first since the Manning's $n$ discussion refers to equations there. 
%    \item Ihab: Our main objective is to show the UQ techniques and how they work. The equations are known and we do not need to stress on them. We can even remove the whole equation or forward model section.
%\end{itemize}
%}

### 2014-01-14 paper/tsunami.tex
fixed figures and removed comments from paper

Manning's $n$ law provides a relationship between the friction coefficient
$C_f$ appearing in the shallow water equations (SWE) governing tsunamis and 
the empirically derived roughness parameter $n$:
\begin{equation}
    C_f = \frac{g n^2}{h^{4/3}},
\label{eq:coef}
\end{equation}
The value of $n$ can vary both spatially and temporally due to changes in the
bottom roughness from wave action and other processes.  For tsunami modeling it
is a common practice to assume a single value of Manning's $n$ for the areas
that are initially below sea-level and sometimes use satellite images and 
land-use data to determine the coefficient in areas above sea-level.  This
process both above and below sea-level has several sources of uncertainty, all
of which can have an impact on the quantities of interest in tsunami modeling.   
In MacInnes \emph{et al.}, for
instance, they found that adopting a uniform value of $n = 0.025$ everywhere
lead to poor performance of the forward model's predicted inundation levels.  To
address this they used $n = 0.035$ in the Sendai plane to account for land-uses 
such as pasture, farmland, and rice paddies.  It is uncertainties such as these 
that we wish to address in a systematic way so that accurate risk assessments 
can be done.

@@

\comment{Kyle - I kind of figured the MacInnes paper goes into detail on this
so I am not sure what pertinent information should be included here.  We could
do some more info on the fault itself and some shore maximum inundations but we
really aren't using any of that.}

@@

The primary computational kernel in \geoclaw is the evaluation of the solution
to the Riemann problem at each grid cell interface.  The Riemann solver used
includes the ability to handle inundation (wet-dry interfaces), well-balancing,
even when the momentum is non-zero, \alert{and entropy corrections}
\cite{George:2008aa}.

One unique feature that \geoclaw has is the ability to adaptively refine the
grids used for the computation essentially following a region of interest as
time progresses (in this case, the disturbance in the surface height $\eta$).
\geoclaw implements these schemes following \cite{Berger:1984ui,Berger:1998aa}
and is described in detail in \cite{Berger:2011du} in the case of tsunami
modeling.

@@

\comment{Still working on getting the Saito \emph{et al.} source model, seems to do better at the DART buoys.}

\comment{We need to address not using the finest bathymetry better}

>>>>

Manning's $n$ law provides a relationship between the friction coefficient
$C_f$ appearing in the shallow water equations (SWE) governing tsunamis and 
the empirically derived roughness parameter $n$:

\begin{equation}
    C_f = \frac{g n^2}{h^{4/3}},
\label{eq:coef}
\end{equation}
The value of $n$ can vary both spatially and temporally due to changes in the
bottom roughness from wave action and other processes.  For tsunami modeling it
is a common practice to assume a single value of Manning's $n$ for the areas
that are initially below sea-level and sometimes use satellite images and 
land-use data to determine the coefficient in areas above sea-level.  This
process both above and below sea-level has several sources of uncertainty, all
of which can have an impact on the quantities of interest in tsunami modeling.   
In MacInnes \emph{et al.}, for
instance, they found that adopting a uniform value of $n = 0.025$ everywhere
lead to poor performance of the forward model's predicted inundation levels.  To
address this they used $n = 0.035$ in the Sendai plane to account for land-uses 
such as pasture, farmland, and rice paddies.  It is uncertainties such as these 
that we wish to address in a systematic way so that accurate risk assessments 
can be done.

@@

%\comment{Kyle - I kind of figured the MacInnes paper goes into detail on this
%so I am not sure what pertinent information should be included here.  We could
%do some more info on the fault itself and some shore maximum inundations but we
%really aren't using any of that.}

@@

The primary computational kernel in \geoclaw is the evaluation of the solution
to the Riemann problem at each grid cell interface.  The Riemann solver used
includes the ability to handle inundation (wet-dry interfaces), well-balancing,
even when the momentum is non-zero, and entropy corrections
\cite{George:2008aa}.

One unique feature that \geoclaw has is the ability to adaptively refine the
grids used for the computation essentially following a region of interest as
time progresses (in this case, the disturbance in the surface height).
\geoclaw implements these schemes following \cite{Berger:1984ui,Berger:1998aa}
and is described in detail in \cite{Berger:2011du} in the case of tsunami
modeling.

@@

%\comment{Still working on getting the Saito \emph{et al.} source model, seems to do better at the DART buoys.}

%\comment{We need to address not using the finest bathymetry better}
